A New Model to Predict Survival in Patients with End-Stage Liver Disease

### 2022-07-17 analysis/article.Rmd
feat: add machine learning to keywords
refactor: integrate Thorsten's review comments

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A New Model to Predict Survival in Patients with End-Stage Liver Disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Objectives**:
      The shortage in grafts for liver transplantation requires risk stratification and allocation rules.
      This study aims to improve the model of end-stage liver disease (MELD) score for 90-day mortality prediction as an risk stratification tool with the help of machine learning (ML) algorithms.

  **Methods**:
      We retrospectively analysed clinical and laboratory data of 654 patients who were recruited during the evaluation process for liver transplantation at the University Hospital Leipzig.
      After comparing 13 different ML algorithms in a nested cross-validation setting and selecting the best performing one we build a new model to predict 90-day mortality in patients with end-stage liver disease.

  **Results**:
      Penalized regression algorithms yield the highest prediction performance in our ML algorithm benchmark.
      In favour of a simpler model we chose the least absolute shrinkage and selection operator (lasso) regression.
      Beside the classical MELD variables INR and bilirubin the lasso regression selected
      cystatin C over creatinine, and additionally IL-6, total protein and cholinesterase.
      The new model offers a better discrimination and calibration than MELD and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.

  **Conclusion**:
      We provide a new model of end-stage liver disease that outperforms the classical MELD score for 90-day survival prediction.

  **Keywords**:
      end-stage liver disease; MELD; risk stratification; survival estimation; IL-6.

@@

Liver cirrhosis is the terminal result of fibrotic remodeling
of liver tissue due to chronic damage. An organ failure is often
irreversible and the only available therapy is liver transplantation.
However, the shortage in grafts for transplantation from deceased donors
requires risk stratification and precise allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates the patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine and international normalized
ratio (INR). In general the MELD score is extended by the sodium level
(MELD-Na score) as this was found to be an important additional risk factor in
liver cirrhosis [@kim2008; @optn_policies2021].
The MELD was initially developed for predicting the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
Afterwards it was revalidated for predicting mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score especially the
creatinine and INR are highly dependent on the utilized laboratory methods
[@trotter2004; @cholongitas2007]. Patients with identical disease state could
have very different MELD scores and thus get different priority on the liver
transplantation waiting list.
Furthermore often, e.g. for acute-on-chronic liver failure, the MELD score
underestimates the mortality risk [@hernaez2020].

@@

AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) across all competitors.
However, the difference against the second best model `r names(roc)[2]`
(AUROC: `r round(roc[2], 3)`) is not statistical significant (
`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`) after correction for multiple
testing (see Table \@ref(tab:roctest) for complete comparison).

@@

rfrank <- as.vector(rank(-rngr$variable.importance, ties = "first")["NA_S"])
rfranktxt <-
    paste0("in a random forest variable importance rank ", rfrank, " of ", ncol(amelddata$x))
```
Although the Organ Procurement and Transplantation Network recommends the use
of MELD-Na over MELD the sodium does not add any information to the score.
Neither in the bootstrap selection nor in the variable importance of a
random forest model the sodium gains any important weight
(`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
Already @kartoun2017 have shown that the
discrimination performance of MELD-Na is insignificantly different from the
classical MELD.
Consequentially it is not a component of our model.

It has shown that women get lower scores in MELD-Na and therefore
are less likely getting an urgent listing for liver transplantation
[@gaertner2022].
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017].
Additionally this transformation yield comparable laboratory measurements
across different clinical laboratories and methods.
It also reduce the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample depending scaling prior to model development.
Both should increase the external applicability of our approach.

@@

Our study has several limitations.
Principally it is a single center, retrospective analysis and without an
external validation cohort.
The data were collected during the evaluation process for liver transplantation
at the University Hospital of Leipzig.
There was no controlled follow-up process. Instead the follow-up information
were recored at the next clinical admission.
Furthermore we excluded all patients who had already a liver transplantation
which could artificial lower the disease severity of the cohort.
Next most of our patients suffered from ethyltoxic liver cirrhosis
(`r .tbl1("Ethyltoxic", pattern = "{p}%")`) which may
limited the transfer to other etiologies.

With a median MELD score of `r .tbl1("ScoreMeldUnos")` our
cohort was healthier than in previous studies that reported
a score of 14-15 [@kim2008; @kartoun2017].
However the 90-day mortality rate of `r .tbl1("M90")` was comparable to
the mortality rates seen in earlier studies that ranges from 6 to 30%
[@malinchoc2000; @kim2008; @kartoun2017].
Although our cohort was healthier with a medium mortality rate the classical
MELD score underestimates the observed mortality
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times.
Unfortunately our sample size is very low in the high risk patient category
which yield a poor but better than MELD based prediction performance of our
AMELD for very ill patients.

@@

The calculations were conducted using the high performance computing (HPC)
cluster of the University Greifswald.
In 2018 the average carbon efficiency of the German power grid was
`r kgco2kwh` kgCO~2~eq/kWh [@icha2021].
A single run of the complete calculation pipeline took roughly
`r round(sum(tm$runtime) / 3600)` hours of computation on an
Intel(R) Xeon(R) E5-2650 v4 and
Intel(R) Xeon(R) Gold 6240 CPU (thermal design power: 105 and 150 W).
The total estimated emissions were at least
`r sprintf("%.2f", kgco2e)` kgCO~2~eq
(ignoring development, test runs, etc.).
This is equivalent to `r round(km)` km driven by an average car
[@allekotte2021].

>>>>

---
# R markdown template taken from
# Dan Holmes <dtholmes@mail.ubc.ca>
# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
title: "A New Model to Predict Survival in Patients with End-Stage Liver Disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Objectives**:
      The shortage in grafts for liver transplantation requires
      risk stratification and adequate allocation rules.
      This study aims to improve the model of end-stage liver disease (MELD)
      score for 90-day mortality prediction with the help of different
      machine learning (ML) algorithms.

  **Methods**:
      We retrospectively analyzed clinical and laboratory data of 654
      patients who were recruited during the evaluation process for
      liver transplantation at the University Hospital Leipzig.
      After comparing 13 different ML algorithms in a nested cross-validation
      setting and selecting the best performing one we build a new model
      to predict 90-day mortality in patients with end-stage liver disease.

  **Results**:
      Penalized regression algorithms yield the highest prediction performance
      in our ML algorithm benchmark.
      In favour of a simpler model we chose the least absolute shrinkage and
      selection operator (lasso) regression.
      Beside the classical MELD variables INR and bilirubin the
      lasso regression selected cystatin C over creatinine, and
      additionally IL-6, total protein and cholinesterase.
      The new model offers an improved discrimination and calibration than MELD
      and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.

  **Conclusion**:
      We provide a new ML based model of end-stage liver disease that
      outperforms the classical MELD score for 90-day survival prediction.

  **Keywords**:
      end-stage liver disease; MELD; risk stratification; machine learning;
      survival estimation; IL-6.

@@

Liver cirrhosis is the terminal result of fibrotic remodeling
of liver tissue due to chronic damage.
This end-stage of liver disease is usually irreversible and
the only available therapy is liver transplantation.
However, the shortage in grafts for transplantation from deceased donors
requires risk stratification and precise and fair allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates the patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine and international normalized
ratio (INR).
The MELD score has been extended by the sodium level
(MELD-Na score) as this was found to be an important additional risk factor in
liver cirrhosis [@kim2008; @optn_policies2021].
The MELD was initially developed for predicting the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
Afterwards it was revalidated for predicting mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score especially the
creatinine and INR are highly dependent on the utilized laboratory methods
[@trotter2004; @cholongitas2007]. Patients with identical disease state could
have very different MELD scores and thus get different priority on the liver
transplantation waiting list.
Furthermore often, e.g. for acute-on-chronic liver failure, the MELD score
underestimates the mortality risk [@hernaez2020].

@@

AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) across all competitors.
However, after correction for multiple testing the difference against
the second best model `r names(roc)[2]`
(AUROC: `r round(roc[2], 3)`) is not statistical significant
(`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
(see Table \@ref(tab:roctest) for complete comparison).

@@

rfrank <- as.vector(rank(-rngr$variable.importance, ties = "first")["NA_S"])
rfranktxt <-
    paste0("in a random forest variable importance rank ", rfrank, " of ", ncol(amelddata$x))
```
Although the Organ Procurement and Transplantation Network recommends the use
of MELD-Na over MELD the sodium does not add any information to the score.
Neither in the bootstrap selection nor in the variable importance of a
random forest model the sodium gains any important weight
(`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
In addition sodium levels can be affected by diuretic therapy and already
@kartoun2017 have shown that the discrimination performance of MELD-Na is
insignificantly different from the classical MELD.
Consequentially it is not a component of our model.

It has shown that women receive in median lower scores in MELD-Na and therefore
are less likely getting an urgent listing for liver transplantation
[@gaertner2022].
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017] in our study.
Additionally this transformation yield comparable laboratory measurements
across different clinical laboratories and methods.
It also reduce the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample depending scaling prior to model development.
Both should increase the external applicability of our approach.

@@

Our study has several limitations.
Principally it is a single center, retrospective analysis and without an
external validation cohort.
The data were collected during the evaluation process for liver transplantation
at the University Hospital of Leipzig.
There was no controlled follow-up process.
We excluded all patients who had already a liver transplantation
which could artificial lower the disease severity of the cohort.
Furthermore most of our patients suffered from ethyltoxic liver cirrhosis
(`r .tbl1("Ethyltoxic", pattern = "{p}%")`) which may
limited the transfer to other etiologies.

With a median MELD score of `r .tbl1("ScoreMeldUnos")` our
cohort was healthier than in previous studies that reported
a score of 14-15 [@kim2008; @kartoun2017].
However the 90-day mortality rate of `r .tbl1("M90")` was comparable to
the mortality rates seen in earlier studies that ranges from 6 to 30%
[@malinchoc2000; @kim2008; @kartoun2017].
Although our cohort was healthier with a medium mortality rate the classical
MELD score underestimates the observed mortality
`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
times.
Unfortunately our sample size is low in the high risk patient category
which yield a poor but better than MELD based prediction performance of our
AMELD for very ill patients.

@@

The calculations were conducted using the high-performance computing (HPC)
cluster of the University Greifswald.
In 2018 the average carbon efficiency of the German power grid was
`r kgco2kwh` kgCO~2~eq/kWh [@icha2021].
A single run of the complete calculation pipeline took roughly
`r round(sum(tm$runtime) / 3600)` hours of computation on an
Intel(R) Xeon(R) E5-2650 v4 and
Intel(R) Xeon(R) Gold 6240 CPU (thermal design power: 105 and 150 W).
The total estimated emissions were at least
`r sprintf("%.2f", kgco2e)` kgCO~2~eq
(ignoring development, test runs, etc.).
This is equivalent to `r round(km)` km driven by an average car
[@allekotte2021].
